Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-08-12 manuscript/main.tex
the function space needs to have second derivatives

\section{Data Objects and Model Formulation} \jongmin{still revising}
For $k = 1, \ldots, K$, 
we assume that the functional predictors $X_{ik}(t)$ are observed on the domain $0 \leq t \leq 1$.and belong to $L^2([0,1])$, the Hilbert space of square-integrable real-valued functions with respect to the Lebesgue measure $dt$ on $[0,1]$.
Additionally, let $\mathbf{Z}=(Z_1, \ldots, Z_p)^\top$
denote a $p$-dimensional multivariate scalar data in $\mathbb{R}^p$, assumed to be a random vector with finite first two moments, equipped with the Euclidean inner product and norm.
 % i.e., for $\mathbf{v_1}=(v_{11}, \ldots, v_{1p})^\top$ and $\mathbf{v_2}=(v_{21}, \ldots, v_{2p})^\top$ in $\mathbb{R}^p$, $\langle \mathbf{v_1}, \mathbf{v_2} \rangle = \mathbf{v_1}^\top \mathbf{v_2} = \sum_{r=1}^p v_{1r}v_{2r}$, and $\Vert \mathbf{v}_1 \Vert = \langle \mathbf{v}_1, \mathbf{v}_1 \rangle^{1/2} = (\sum_{r=1}^p v_{1r}^2)^{1/2}$.
The hybrid random predictor and corresponding inner product is formally defined as follows.
\begin{definition}[Hybrid predictor and inner product]\label{def:hilbert_space}
	We define a  hybrid predictor  $W_i$ as an ordered pair combining the functional predictors $X_{i1}(t), \ldots, X_{iK}(t)$ and a scalar predictor $\mathbf{Z}_i$:
\begin{equation}\label{def:hybid_predictor}
		W_i := (X_{i1}, \ldots, X_{iK}, \mathbf{Z}_i)
		\in \mathcal{H} :=  L^2([0,1])^{\otimes K} \times \mathbb{R}^p
\end{equation}
	The inner product between two hybrid predictors,
	$W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$
	and
	$W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, is then defined as:
	\begin{equation}		\label{eq: hybrid inner product}
		\langle W_1, W_2\rangle_{\mathcal{H}}
		:=
		\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
		+
		\omega
		\mathbf{Z}_1^\top \mathbf{Z}_2,
	\end{equation}
	for $\omega > 0$.
	The corresponding norm $\Vert \cdot \Vert_\mathcal{H}$ is derived from this inner product as $\langle \cdot,\cdot \rangle_\mathcal{H}^{1/2}$.
\end{definition}
In \eqref{eq: hybrid inner product}, 
$\omega$ is a positive weight that needs to be pre-specified or estimated. It is mainly used to take into account heterogeneity between functional and scalar parts in terms of measurement scale and/or amount of variation (see Section \ref{subsec: Data Preprocessing}).

@@

\begin{definition}[Roughness-sensitive inner product]
	$\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}$
\end{definition}
  ($\widehat{\xi}_1, \widehat{\xi}_2, \ldots$)  based on the modified inner product  which now takes into account the roughness of the functions:
\begin{align*}
	\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
	&= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right \rangle_{L^2} \\
	&= \sum \limits_{k=1}^K \int_0^1 \widehat{\psi}_l^{(k)}(t_k) \widehat{\psi}_j^{(k)}(t_k) dt_k + \sum \limits_{r=1}^p \widehat{\theta}_{lr} \widehat{\theta}_{jr} + \sum_{k=1}^K \lambda_k \int_0^1 \ddot{\widehat{\psi}}_l^{(k)}(t_k) \ddot{\widehat{\psi}}_j^{(k)}(t_k) dt_k \\
	& = \mathbb{1}(l=j). 
\end{align*}
This modified orthonormality condition is one of the main geometric properties of the proposed PLS algorithm and is formally stated with proof in Section \ref{section:main:properties_of_our_method} (see Proposition \ref{proposition: modified orthnormality of PLS components}), along with other properties.

>>>>

\section{Data Objects and Model Formulation} \jongmin{still revising}
For $k = 1, \ldots, K$, 
we assume that the functional predictors $X_{ik}(t)$ are observed on the domain $0 \leq t \leq 1$, and belong to $\mathcal{F}$,
which is defined as the space of real-valued functions on $[0,1]$
which are square-integrable with respect to the Lebesgue measure $dt$ on $[0,1]$,
 and possess second  derivatives.
Additionally, let $\mathbf{Z}=(Z_1, \ldots, Z_p)^\top$
denote a $p$-dimensional multivariate scalar data in $\mathbb{R}^p$, assumed to be a random vector with finite first two moments, equipped with the Euclidean inner product and norm.
 % i.e., for $\mathbf{v_1}=(v_{11}, \ldots, v_{1p})^\top$ and $\mathbf{v_2}=(v_{21}, \ldots, v_{2p})^\top$ in $\mathbb{R}^p$, $\langle \mathbf{v_1}, \mathbf{v_2} \rangle = \mathbf{v_1}^\top \mathbf{v_2} = \sum_{r=1}^p v_{1r}v_{2r}$, and $\Vert \mathbf{v}_1 \Vert = \langle \mathbf{v}_1, \mathbf{v}_1 \rangle^{1/2} = (\sum_{r=1}^p v_{1r}^2)^{1/2}$.
The hybrid random predictor and corresponding inner product is formally defined as follows.
\begin{definition}[Hybrid predictor and inner product]\label{def:hilbert_space}
	We define a  hybrid predictor  $W_i$ as an ordered pair combining the functional predictors $X_{i1}(t), \ldots, X_{iK}(t)$ and a scalar predictor $\mathbf{Z}_i$:
\begin{equation}\label{def:hybid_predictor}
		W_i := (X_{i1}, \ldots, X_{iK}, \mathbf{Z}_i)
		\in \mathcal{H} :=  \mathcal{F}^{\otimes K} \times \mathbb{R}^p
\end{equation}
	The inner product between two hybrid predictors,
	$W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$
	and
	$W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$, is then defined as:
	\begin{equation}		\label{eq: hybrid inner product}
		\langle W_1, W_2\rangle_{\mathcal{H}}
		:=
		\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
		+ 
		\omega
		\mathbf{Z}_1^\top \mathbf{Z}_2,
	\end{equation}
	for $\omega > 0$.
	The corresponding norm $\Vert \cdot \Vert_\mathcal{H}$ is derived from this inner product as $\langle \cdot,\cdot \rangle_\mathcal{H}^{1/2}$.
\end{definition}
In \eqref{eq: hybrid inner product}, 
$\omega$ is a positive weight that needs to be pre-specified or estimated. It is mainly used to take into account heterogeneity between functional and scalar parts in terms of measurement scale and/or amount of variation (see Section \ref{subsec: Data Preprocessing}).

@@

\begin{definition}[Roughness-sensitive inner product]
		Given two hybrid predictors 	$W_1 = (X_{11}, \ldots, X_{1K}, \mathbf{Z}_1)$
	and
	$W_2 = (X_{21}, \ldots, X_{2K}, \mathbf{Z}_2)$ belonging to $\mathcal{H}$ defined in Definition \ref{def:hybid_predictor}
	and roughness penalty matrix 	$\Lambda  = \operatorname{blkdiag}(\lambda_1 I_M, \ldots, \lambda_K I_M, 0_{p \times p})$,
	The inner product between two hybrid predictors, taking into the roughness peanlties,  is then defined as:
\begin{equation}
	\langle \langle h_l, h_j \rangle \rangle = \langle h_l, h_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \langle \ddot{f}_l^{(k)}, \ddot{f}_j^{(k)} \rangle_{L^2}
\end{equation}
	\begin{equation}		\label{eq: hybrid inner product}
	\langle W_1, W_2\rangle_{\mathcal{H}, \Lambda}
	:=
	\sum \limits_{k=1}^K \int_0^1 X_{1k}(t) X_{2k}(t) \, dt
	\sum \limits_{k=1}^K \int_0^1 X^{\prime \prime}_{1k}(t) X^{\prime \prime}_{2k}(t) \, dt
	+ 
	\omega
	\mathbf{Z}_1^\top \mathbf{Z}_2,
\end{equation}
for $\omega > 0$.
The corresponding norm $\Vert \cdot \Vert_\mathcal{H}$ is derived from this inner product as $\langle \cdot,\cdot \rangle_\mathcal{H}^{1/2}$.
\end{definition}

  ($\widehat{\xi}_1, \widehat{\xi}_2, \ldots$)  based on the modified inner product  which now takes into account the roughness of the functions:
\begin{align*}
	\langle \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle \rangle 
	&= \langle \widehat{\xi}_l, \widehat{\xi}_j \rangle_{\mathcal{H}} + \sum_{k=1}^K \lambda_k \left\langle \ddot{\widehat{\psi}}_l^{(k)}, \ddot{\widehat{\psi}}_j^{(k)} \right \rangle_{L^2} \\
	&= \sum \limits_{k=1}^K \int_0^1 \widehat{\psi}_l^{(k)}(t_k) \widehat{\psi}_j^{(k)}(t_k) dt_k + \sum \limits_{r=1}^p \widehat{\theta}_{lr} \widehat{\theta}_{jr} + \sum_{k=1}^K \lambda_k \int_0^1 \ddot{\widehat{\psi}}_l^{(k)}(t_k) \ddot{\widehat{\psi}}_j^{(k)}(t_k) dt_k \\
	& = \mathbb{1}(l=j). 
\end{align*}
This modified orthonormality condition is one of the main geometric properties of the proposed PLS algorithm and is formally stated with proof in Section \ref{section:main:properties_of_our_method} (see Proposition \ref{proposition: modified orthnormality of PLS components}), along with other properties.
